Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-23 paper/paper.tex
Use singular form of control

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
\documentclass[acmsmall]{acmart}\settopmatter{}

@@

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract  definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and un-CPS to direct-style with delimited controls. Linearization expresses
  non-deterministic choices by first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the nondeterminism 
  of abstract interpretation.

@@

\begin{itemize}
  \item We show \textbf{linearization} in Section~\ref{linear}. By expressing the
    nondeterministic choices as a first-order data type, we linearize the execution
    of abstract abstract machines; the transition of machine states therefore becomes 
    deterministic. Notably, this introduces another layer of controls that will also 
    be refunctionalized later.

@@

  \item The last transformation we show is that of transforming the refunctionalized AAM to a
    \textbf{direct-style} interpreter (Section~\ref{directstyle}) by using delimited controls.
\end{itemize}

@@

% \note{TR: can this be merged with contributions? (they currently lack section refs)}
% The remainder of this paper is organized as follows.
% We begin by describing the target language we will analyze in Section~\ref{background},
% including a concrete CESK machine as small-step operational semantics for it.
% We continue with a review of abstracting abstract machines (AAM) as an abstract interpretation
% counterpart of CESK machines. Following this, Section~\ref{linear} shows a conversion from the
% nondeterministic choices of closures to another layer of continuations,
% therefore linearizing the abstract execution.
% Section~\ref{fusing} then discusses fusing the driver and step functions to a single
% function.
% We further identify in Section~\ref{disen} the first-order data type representations of
% continuations, and disentangle the code blocks for dispatching data types to separate
% functions.
% Section~\ref{refunc} then represents the refunctionalized abstract abstract machines.
% In this section, the continuations represented by data types become higher-order
% functions.
% In Section~\ref{directstyle}, we transform the refunctionalized AAM to two
% different direct styles: one implemented with monads, the other using
% delimited controls.

@@

As a consequence, the components of states are also changed: the store is divided
into binding stores and continuation stores; the continuation stack becomes an address.
By dereferencing this address in a continuation store, we can retrieve the actual 
transfer of controls. The definition of environment @Env@ remains the same.

@@

In order to address this problem, our first step of transformations
is to linearize all nondeterministic choices.
This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of controls, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state, which we now call @NDState@.
For clarity of presentation in differentiating between the two continuations, we elect to call
the first a \emph{normal continuation}, and the second a \emph{nondeterministic continuation}.

@@

These coincidences should not be a surprise since existing literature is extraordinarily rich
in showing the correspondence between monads and continuation-passing style as well as
delimited controls since their very beginning era \cite{Danvy:1990:AC:91556.91622, wadler1992essence,
danvy1992representing, moggi1991notions}.

@@

\subsection{Un-CPS by Delimited Controls} \label{uncps}

As previously identified, there are two layers of continuations: one for the normal stack 
of the analyzed language, and the other for the nondeterministic choices of closures.
This corresponds to the extended continuation-passing style.
But in our defining language (Scala), the delimited controls rely on the continuation-passing 
transformation which is only one layer.

@@

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus,
instead of plain $\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.
But instead of directly showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with
delimited controls. 
We use the delimited control operators @shift@ and @reset@ that are implemented in Scala 
\cite{rompf2009implementing}. The correspondence of delimited controls (as well as continuations) 
and monads are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
\documentclass[acmsmall, review]{acmart}\settopmatter{}

@@

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract  definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and un-CPS to direct-style with delimited control. Linearization expresses
  non-deterministic choices by first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the nondeterminism 
  of abstract interpretation.

@@

\begin{itemize}
  \item We show \textbf{linearization} in Section~\ref{linear}. By expressing the
    nondeterministic choices as a first-order data type, we linearize the execution
    of abstract abstract machines; the transition of machine states therefore becomes 
    deterministic. Notably, this introduces another layer of control that will also 
    be refunctionalized later.

@@

  \item The last transformation we show is that of transforming the refunctionalized AAM to a
    \textbf{direct-style} interpreter (Section~\ref{directstyle}) by using delimited control operators.
\end{itemize}

@@

% \note{TR: can this be merged with contributions? (they currently lack section refs)}
% The remainder of this paper is organized as follows.
% We begin by describing the target language we will analyze in Section~\ref{background},
% including a concrete CESK machine as small-step operational semantics for it.
% We continue with a review of abstracting abstract machines (AAM) as an abstract interpretation
% counterpart of CESK machines. Following this, Section~\ref{linear} shows a conversion from the
% nondeterministic choices of closures to another layer of continuations,
% therefore linearizing the abstract execution.
% Section~\ref{fusing} then discusses fusing the driver and step functions to a single
% function.
% We further identify in Section~\ref{disen} the first-order data type representations of
% continuations, and disentangle the code blocks for dispatching data types to separate
% functions.
% Section~\ref{refunc} then represents the refunctionalized abstract abstract machines.
% In this section, the continuations represented by data types become higher-order
% functions.
% In Section~\ref{directstyle}, we transform the refunctionalized AAM to two
% different direct styles: one implemented with monads, the other using
% delimited control operators.

@@

As a consequence, the components of states are also changed: the store is divided
into binding stores and continuation stores; the continuation stack becomes an address.
By dereferencing this address in a continuation store, we can retrieve the actual 
transfer of control. The definition of environment @Env@ remains the same.

@@

In order to address this problem, our first step of transformations
is to linearize all nondeterministic choices.
This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of control, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state, which we now call @NDState@.
For clarity of presentation in differentiating between the two continuations, we elect to call
the first a \emph{normal continuation}, and the second a \emph{nondeterministic continuation}.

@@

These coincidences should not be a surprise since existing literature is extraordinarily rich
in showing the correspondence between monads and continuation-passing style as well as
delimited control since their very beginning era \cite{Danvy:1990:AC:91556.91622, wadler1992essence,
danvy1992representing, moggi1991notions}.

@@

\subsection{Back to Direct-Style with Delimited Control} \label{uncps}

As previously identified, there are two layers of continuations: one for the normal stack 
of the analyzed language, and the other for the nondeterministic choices of closures.
This corresponds to the extended continuation-passing style.
But in our defining language (Scala), the delimited control operators rely on the continuation-passing 
transformation which is only one layer.

@@

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus,
instead of plain $\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.
But instead of directly showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with
delimited control operators. 
We use the delimited control operators @shift@ and @reset@ that are implemented in Scala 
\cite{rompf2009implementing}. The correspondence of delimited control (as well as continuations) 
and monads are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.
